Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-03-31 article/01_Introduction.Rmd
phrasing

The classic narrative of the development of these rites in Europe presented e.g. by @hardingEuropeanSocietiesBronze2000a (following @hauslerBestattungssittenFruhenBronzezeit1977; @hauslerGrabUndBestattungssitten1994; @hauslerTotenorientierungUndGeographischer1996) can be summarised as follows: In the early and middle Bronze Age inhumations dominated in different variations. Cremation burials in this time only occur in the Hungarian lowlands. Burial mounds are concentrated in parts of the Balkans as well as in Eastern, Western and Northern Europe, while in Central and Southern Europe flat graves -- mostly inhumations -- predominate. In the Middle Bronze Age, barrows gained importance in Western and Central Europe. This observation is highly linked to the emergence of the *Tumulus culture* in Central Europe. In the late Bronze Age, the Urnfield period, cremation became the most common burial custom in the context of *Urnfield culture*. Harding's narrative is, even when taking into account its enormous simplification, partially misleading. Cremation, for example, was a lot more common already in the early and middle Bronze Age, though not in the context of the later Urnfield culture (e.g. in the Benelux area [@fokkensBronzeAgeLow2013] or Great-Britain [@robertsBritainIrelandBronze2013]).

@@

The diffusion of the cremation burial tradition in the middle and late Bronze Age with radiocarbon data was already the subject of promising research in Central and South-Western Europe with focus on Catalonia and Switzerland [@capuzzoCulturalChangesSecond2015; @capuzzoInhumacionCremacionNordeste2017; @capuzzoSpacetemporalAnalysisRadiocarbon2014]. This approach based on the dedicated radiocarbon database EUBAR [@capuzzoEUBARDatabase14C2013] sets a precedent for a more reproducible and statistically sound research practice concerning diffusion of innovation in Bronze Age archaeology.

@@

Burial traditions are a ritual behaviour that is transmitted between generations. They are therefore suitable for study using methods of *Cultural transmission theory*, a special aspect of *Cultural evolution theory* concerned with the spread of ideas in human populations [@creanzaCulturalEvolutionaryTheory2017]. This theoretical framework permits the comparison of behaviour across spatial, temporal and cultural boundaries, as it assumes *Social learning* to produce adaptive behaviour through transmission, selection and mutation -- a secondary heredity system in humans [@henrichDualinheritanceTheoryEvolution2007; @lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002].

*Evolutionary archaeology*, the branch of archaeology that explicitly adopted the terminology of Cultural evolution, established an important distinction of cultural behaviour in *function* and *style* [@dunnell1978style] which is very relevant for burial traditions: Some types of human behaviour increase well-being and simplify survival, like specific hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism, their prevalence generally happens to increase over time. On the other hand, some behaviours are not related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewellery design. Despite or because their seemingly subordinate importance, these characteristics often pass through a breathtakingly fast development. Burial traditions can be described as a special form of these *stylistic traits*. Special because of their strong dependence on religious and spiritual beliefs and their connection with the emotionally stirring topic of death, but still subject to sometimes rapid and seemingly random fashion changes.

Of course in reality most behaviour is not unambiguously stylistic or functional. Burial rites cause costs regarding working time and material to a varying extent. These expenses might have been influenced by or had their effect on the human-nature relation: Pyres, for example, require a significant amount of wood and the erection of barrows much workforce. Still, it can be expected that these functional disadvantages were not critical most of the time -- at least regarding the general decision which kind of burial tradition is applied. Most probably they influence the magnitude of invested energy for a burial, just like the level of grief [@macdonaldGriefBurialAmerican2001] or the representational intentions of the burying relatives [@henrichEvolutionCostlyDisplays2009].

@dunnell1978style postulates that the distribution and emergence of stylistic cultural traits behave randomly on a population scale level. The stochastic processes that are at work are an important object of investigation and discussed as *Neutral Transmission* or *Stylistic variability* [@bentleyRandomDriftCulture2004; @lipoNeutralityStyleDrift2001; @neimanStylisticVariationEvolutionary1995; @porcicSimulatingCulturalTransmission2014; @porcicExploringEffectsAssemblage2015; @rogersNaturalSelectionCultural2008; @shennanCeramicStyleChange2001]. @neimanStylisticVariationEvolutionary1995 takes a closer look at these processes and develops a simple framework for the simulation of non-functional idea propagation. He identifies the essential phenomena *drift* (random dominance of similar traits) and *flow* (intergroup transmission) in population genetics to be also relevant for cultural traits in archaeology. Furthermore, he implements a measure of cultural distance to apply his theoretical insights to real-world data. This workflow was an important methodological inspiration for the burial rites data examination in this paper.

>>>>

The classic narrative of the development of these rites in Europe presented e.g. by @hardingEuropeanSocietiesBronze2000a following @hauslerBestattungssittenFruhenBronzezeit1977; @hauslerGrabUndBestattungssitten1994 and @hauslerTotenorientierungUndGeographischer1996 can be summarised as follows: In the Early and Middle Bronze Age inhumations dominated in different variations. Cremation burials in this time only occurred in the Hungarian lowlands. Burial mounds were concentrated in parts of the Balkans as well as in Eastern, Western and Northern Europe, while in Central and Southern Europe flat graves -- mostly inhumations -- predominated. In the Middle Bronze Age, barrows gained importance in Western and Central Europe. This observation is highly linked to the emergence of the *Tumulus culture* in Central Europe. In the Late Bronze Age, the Urnfield period, cremation became the most common burial custom in the context of *Urnfield culture*. Harding's narrative is, even when taking into account its enormous simplification, partially misleading. Cremation, for example, was a lot more common already in the Early and Middle Bronze Age, though not in the context of the later Urnfield culture (e.g. in the Benelux area [@fokkensBronzeAgeLow2013] or Great-Britain [@robertsBritainIrelandBronze2013]).

@@

Similar methodology was already the subject of promising, but only regional research in Central and South-Western Europe with focus on Catalonia and Switzerland [@capuzzoCulturalChangesSecond2015; @capuzzoInhumacionCremacionNordeste2017; @capuzzoSpacetemporalAnalysisRadiocarbon2014]. This approach based on the dedicated radiocarbon database EUBAR [@capuzzoEUBARDatabase14C2013] aimed to explore the diffusion of the cremation burial tradition in the Middle and Late Bronze Age and set a precedent for a more reproducible and statistically sound research practice in Bronze Age archaeology.

@@

Burial traditions are a ritual behaviour that is transmitted between generations. They are therefore suitable for study using methods of *Cultural transmission theory*, a special aspect of *Cultural evolution theory* concerned with the spread of ideas in human populations [@creanzaCulturalEvolutionaryTheory2017]. This theoretical framework permits the comparison of behaviour across spatial, temporal and cultural boundaries, because it assumes a secondary heredity system in humans: *Social learning* with transmission, selection and mutation of ideas [@henrichDualinheritanceTheoryEvolution2007; @lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002].

*Evolutionary archaeology*, the branch of archaeology that explicitly adopted Cultural evolution theory, established an important distinction of cultural behaviour in *function* and *style* [@dunnell1978style] which is relevant for burial traditions. Some types of human behaviour increase well-being and simplify survival, like specific hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism, their prevalence generally happens to increase over time. On the other hand, some behaviours are not related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewellery design. Despite or because their seemingly subordinate importance, these characteristics often pass through a breathtakingly fast development. Burial traditions can be described as a special form of these *stylistic traits*. Special because of their strong dependence on religious and spiritual beliefs and their connection with the emotionally stirring topic of death, but still subject to sometimes rapid and seemingly random fashion changes.

Of course in reality most behaviour is not unambiguously stylistic or functional. Burial rites cause costs regarding working time and material to a varying extent. These expenses might have been influenced by or had their effect on the human-nature relation: Pyres, for example, require a significant amount of wood and the erection of barrows much workforce. Still, it can be expected that these functional disadvantages were not critical most of the time for burial traditions -- at least regarding the general decision which ritual is performed. Most probably they influenced the magnitude of invested energy for a burial, just like the level of grief [@macdonaldGriefBurialAmerican2001] or the representational intentions of the burying relatives [@henrichEvolutionCostlyDisplays2009].

@dunnell1978style postulates that the distribution and emergence of stylistic cultural traits behave randomly on a population scale level. The stochastic processes that are at work are an important object of investigation and discussed as *Neutral Transmission* or *Stylistic variability* [@bentleyRandomDriftCulture2004; @lipoNeutralityStyleDrift2001; @neimanStylisticVariationEvolutionary1995; @porcicSimulatingCulturalTransmission2014; @porcicExploringEffectsAssemblage2015; @rogersNaturalSelectionCultural2008; @shennanCeramicStyleChange2001]. @neimanStylisticVariationEvolutionary1995 takes a closer look at these processes and develops a simple framework for the simulation of non-functional idea propagation. He identifies the essential phenomena *drift* (i.e. in case of multiple, equivalent traits in a population, a single one will dominate over time due to random effects) and *flow* (i.e. intergroup transmission) in population genetics to be also relevant for cultural traits in archaeology. Furthermore, he implements a measure of cultural distance to apply his theoretical insights to real-world data. This workflow was an important methodological inspiration for the burial rites data examination in this paper.
